# Visualising Time Series Data

```{r}
library(reticulate)
use_python("C:/Users/DELL/AppData/Local/Programs/Python/Python313/python.exe", required = TRUE)

```

## Time series

A time series is a sequence of observations recorded in time order. The time intervals between observations can be regular (e.g., daily, monthly, yearly) or irregular (e.g., magnitude of a earthquake at a particular location).

## Frequency of a time series (Seasonal periods)

Number of observations per natural time interval (Usually year, but sometimes a week, a day, an hour)

### Single Seasonality

The time series exhibits one repeating pattern at a fixed frequency.

Example:

Monthly sales that peak every December (annual seasonality).

```{r}
# Create the dataframe
data_frequency <- data.frame(
  Data = c("Annual", "Quarterly", "Monthly", "Weekly"),
  Frequency = c(1, 4, 12, 52)
)

# View the dataframe
data_frequency |> knitr::kable()
```

### Multiple Seasonality

The time series exhibits more than one repeating pattern at different frequencies simultaneously.

Example:

Hourly electricity demand with a daily pattern (peaks every day at certain hours), a weekly pattern (weekdays vs weekends).

Website traffic with hourly variation and seasonal holiday peaks.

```{r}
# Load knitr
library(knitr)

# Create the dataframe
frequencies <- data.frame(
  Data = c("Daily", "Hourly", "Half-hourly", "Minutes", "Seconds"),
  Minute = c(NA, NA, NA, 60, 60),
  Hour = c(NA, NA, NA, 1440, 3600),
  Day = c(NA, 24, 48, 1440, 86400),
  Week = c(7, 168, 336, 10080, 604800),
  Year = c(365.25, 8766, 17532, 525960, 31557600)
)

# Display as kable
kable(frequencies, align = "c", caption = "Time Unit Frequencies")

```

## `DataFrame` for time series data

When your DataFrame represents a time series, the index is usually the date or time, allowing pandas to:

 - Plot time series easily

 - Resample or aggregate data by time

 - Compute rolling statistics

```{python}
#| echo: true
# Import pandas
#py -m pip install pandas
import pandas as pd

# Define data
value = [100, 250, 78, 300, 500]
time = list(range(2015, 2020))

# Create DataFrame
df = pd.DataFrame({"Year": time, "Observation": value})

# Set 'Year' as index
df.set_index("Year", inplace=True)

# Display the DataFrame
print(df)

```

For data collected more often than once a year (e.g., monthly, weekly, or daily), it’s important to tell the computer that the index represents time. We do this by converting the index to a time or date type using a time-class function. This helps us sort, select, and analyze the data correctly over time.

```{python}
#| echo: true
# Sample monthly data
data = {
    "Month": pd.date_range(start="2025-01-01", periods=6, freq="M"),  # 6 months
    "Sales": [120, 150, 170, 130, 180, 200]
}

# Create DataFrame
z = pd.DataFrame(data)

# Format Month as "Year Month" (e.g., "2025 Jan")
z["Month"] = z["Month"].dt.strftime("%Y %b")

# Set Month as index
z.set_index("Month", inplace=True)

# Display the DataFrame
print(z)

```

## Time series visualisation using grammar of graphics

### Introduction to "grammar of graphics" with plotnine



